Outline

### 2023-05-02 manuscript/manuscript.Rmd
Adding regularized regression discussion and clockplot code

All analyses were run in R, version 4.2.2. Logistic regressions were performed using base R's `glm` function with elastic net functionality provided by the glmnet package, random forest classifiers were performed using the randomForest package, and initial clustering approaches were explored with the dbscan package.

@@

@@

@@

@@

gplist <- c("all_params", "xcms_params", "raw_data_params") %>%
  lapply(function(which_params){
    ggplot() +
      geom_hline(yintercept = 0) +
      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error*2, 
                    ymax=estimate+std.error*2),
                fill="grey80", data=best_estimates %>% filter(param_name==which_params)) +
      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error, 
                    ymax=estimate+std.error),
                fill="grey50", data=best_estimates %>% filter(param_name==which_params)) +
      geom_point(aes(x=sample_fraction, y=estimate), 
                 data=sample_model_estimates %>% filter(param_name==which_params),
                 alpha=0.5) +
      ggh4x::facet_grid2(param_name~term, scales = "free_y", independent = "y") +
      scale_x_continuous(breaks = c(0, 0.2, 0.5, 0.8, 1), limits = c(0, 1)) +
      theme_bw()
  })
sampled_model_est_gp <- do.call(what = egg::ggarrange, c(gplist, ncol=1))
ggsave(plot = sampled_model_est_gp, 
       filename = "sampled_model_est.png", 
       device = "png", path = "manuscript/supplement",
       width = 6.5, height = 5, units = "in", dpi = 300)
```

@@

This stricter threshold was used

```{r FDR vs GPF with additional datasets}
# Currently a fairly simple figure, may need additional data?
both_min_model <- rbind(FT2040_features, MS3000_features) %>%
  select(feat_class, med_cor, med_SNR) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)
pred_probs_all <- bind_rows(lst(
  Falkor=FT2040_features, MESOSCOPE=MS3000_features,
  CultureData=CultureData_features, Pttime=Pttime_features), 
  .id = "dataset") %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.9, "Good", "Bad")) %>%
  mutate(error_type=case_when(
    feat_class=="Good" & pred_class=="Good" ~ "%TP",
    feat_class=="Good" & pred_class=="Bad" ~ "%FN",
    feat_class=="Bad" & pred_class=="Good" ~ "%FP",
    feat_class=="Bad" & pred_class=="Bad" ~ "%TN",
    TRUE ~ "Incomparable"
  )) %>%
  mutate(dataset=factor(dataset, levels=c("MESOSCOPE", "Falkor", "CultureData", "Pttime"))) %>%
  filter(error_type!="Incomparable") %>%
  group_by(dataset, error_type) %>%
  summarise(n=n(), .groups = 'drop') %>%
  complete(dataset, error_type, fill=list(n=0))
error_type_all_datasets <- pred_probs_all %>%
  pivot_wider(names_from=error_type, values_from = n) %>%
  mutate(`FDR`=`%FP`/(`%FP`+`%TP`)) %>%
  mutate(`GPF`=`%TP`/(`%FN`+`%TP`)) %>%
  mutate(`GPF`=ifelse(dataset%in%c("CultureData", "Pttime"), NA, `GPF`)) %>%
  pivot_longer(c(`FDR`, `GPF`)) %>%
  drop_na()

@@

>>>>

We also tested whether the performance could be improved with elastic net regression or random forest models. These models handle correlated variables better than ordinary least squares regression, so we also included several additional implementations of the peak shape and novel SNR parameters when summarizing across multiple files, using a max and a median of the top-three best values rather than just the overall median as well as a log-transformed version of the median peak shape calculated as $median(log_{10}(1-r))$ where $r$ is Pearson's correlation coefficient, as described above (Figure 2). Cross-validation was used to select the optimal tuning parameter $\lambda$ with `glmnet` package's `cv.glmnet` for an elastic net penalty (α) of 0, 0.5, and 1. Random forests were implemented using the randomForest package with default settings and a factor-type response vector to ensure classification was used rather than regression.

We explored the implications of applying this model to the MESOSCOPE dataset at a variety of thresholds. In univariate space, we used nonparametric Kruskal-Wallis analyses of variance to measure the difference between the surface (15m), DCM (~110m), and 175m samples because the metabolite peak areas could not be assumed to be normally distributed. These univariate tests were then controlled for multiply hypothesis testing using R's `p.adjust` function with method `fdr`.[@ref] We also performed post-hoc Dunn tests provided by the `rstatix` package to categorize the response to depth for those mass features for which the KW test was significant, with responses falling into one of the 14 classes possible when permuting the sign and significance of the Dunn test.[@ref] p-values obtained from the Dunn tests were not FDR controlled because it was used as a categorization tool rather than a null hypothesis test. In multivariate space, we used a permutational MANOVA (PERMANOVA) [@ref] provided by the vegan package's `adonis2` function to test for multivariate differences in structure of the metabolome with depth.[@ref] We ran multiple PERMANOVAs with a different subset of mass features included each time, corresponding to using the output from XCMS directly, likelihood thresholds of 0.01, 0.1, 0.5, 0.9, and finally only those peaks manually annotated as good.

All analyses were run in R, version 4.2.2, and code is available on the Github at https://github.com/wkumler/MS_metrics.[@ref]

@@

### Dataset characterization

@@

### Logistic regression performance

@@

### Model stability under different training sets and subset of labels

@@

gplist <- c("all_params", "xcms_params", "raw_data_params") %>%
  lapply(function(which_params){
    ggplot() +
      geom_hline(yintercept = 0) +
      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error*2, 
                    ymax=estimate+std.error*2),
                fill="grey80", data=best_estimates %>% filter(param_name==which_params)) +
      geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=estimate-std.error, 
                    ymax=estimate+std.error),
                fill="grey50", data=best_estimates %>% filter(param_name==which_params)) +
      geom_point(aes(x=sample_fraction, y=estimate), 
                 data=sample_model_estimates %>% filter(param_name==which_params),
                 alpha=0.5) +
      ggh4x::facet_grid2(param_name~term, scales = "free_y", independent = "y") +
      scale_x_continuous(breaks = c(0, 0.2, 0.5, 0.8, 1), limits = c(0, 1)) +
      theme_bw()
  })
sampled_model_est_gp <- do.call(what = egg::ggarrange, c(gplist, ncol=1, draw=FALSE))
ggsave(plot = sampled_model_est_gp, 
       filename = "sampled_model_est.png", 
       device = "png", path = "manuscript/supplement",
       width = 6.5, height = 5, units = "in", dpi = 300)
```

### Performance of a stricter threshold on novel datasets

@@

This stricter threshold was chosen because we felt less comfortable interpreting results based on mass features that were only 50% likely to be real, but did not feel the need to be so strict with this exploratory analysis that we wanted to limit it to 99+% likelihood peaks. This threshold also struck a nice balance between the number of peaks we estimated to be necessary for robust testing while still remaining reasonable to manually label. For the CultureData dataset, we obtained 1,790 total mass features, 192 of which had predicted likelihoods above 0.9. Of these, 151 were identified manually as "Good", 21 were given "Ambiguous" designations, and only 3 were flagged as "Bad", with the remaining 17 appearing only in the standards. For the Pttime dataset, 7,781 were obtained with 400 flagged by the model as "Good". 348 were truly good peaks, 35 were ambiguous, and 17 were "Bad". No standards were run during this analysis, so there were no features in that category. [There exists, however, some uncontrolled confirmation bias here in the manual labelling because the MS expert knew that these peaks had already passed some quality threshold and were more likely to be good than not.]

With the stricter threshold, we obtained FDR rates consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for the Falkor, MESOSCOPE, CultureData, and Pttime respectively [Figure 6]. Of course, this low error rate required that we miss out on additional valuable features, with only a fraction of the total good peaks making it past this threshold. In both the Falkor and MESOSCOPE datasets, the fraction of good peaks found was below 40% with actual values of 39.4% and 26.5%, respectively. Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the %GPF but expect it to be in a similar range (Figure 6).

```{r FDR vs GPF with additional datasets}
# Currently a fairly simple figure, may need additional data?
both_min_model <- rbind(FT2040_features, MS3000_features) %>%
  select(feat_class, med_cor, med_SNR) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)
pred_probs_all <- bind_rows(lst(
  Falkor=FT2040_features, MESOSCOPE=MS3000_features,
  CultureData=CultureData_features, Pttime=Pttime_features), 
  .id = "dataset") %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.9, "Good", "Bad")) %>%
  # count(pred_class, dataset)
  mutate(error_type=case_when(
    feat_class=="Good" & pred_class=="Good" ~ "%TP",
    feat_class=="Good" & pred_class=="Bad" ~ "%FN",
    feat_class=="Bad" & pred_class=="Good" ~ "%FP",
    feat_class=="Bad" & pred_class=="Bad" ~ "%TN",
    TRUE ~ "Incomparable"
  )) %>%
  # count(feat_class, pred_class, dataset) %>% filter(dataset=="Pttime")
  mutate(dataset=factor(dataset, levels=c("MESOSCOPE", "Falkor", "CultureData", "Pttime"))) %>%
  filter(error_type!="Incomparable") %>%
  group_by(dataset, error_type) %>%
  summarise(n=n(), .groups = 'drop') %>%
  complete(dataset, error_type, fill=list(n=0))
error_type_all_datasets <- pred_probs_all %>%
  pivot_wider(names_from=error_type, values_from = n) %>%
  mutate(`FDR`=`%FP`/(`%FP`+`%TP`)) %>%
  mutate(`GPF`=`%TP`/(`%FN`+`%TP`)) %>%
  mutate(`GPF`=ifelse(dataset%in%c("CultureData", "Pttime"), NA, `GPF`)) %>%
  pivot_longer(c(`FDR`, `GPF`)) %>%
  drop_na()

@@

### Regularized regression and random forests perform about the same

None of the advanced ML models significantly improved performance when cross-validated between the MESOSCOPE and Falkor datasets when measured by both initial performance and the performance drop when applied across datasets. All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GPF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent Type II error acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp. figure X). The random forest model had perfect predictive capacity when tested internally on the training data (FDR=0%, GPF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp. figure X). In each case, the performance drop when applied to a novel dataset was more extreme than the simple two-parameter model described above.

```{r elastic net and random forest}
library(glmnet)
net_cross_train_df <- lst(ridge=0, elastic=0.5, lasso=1) %>%
  imap(function(alpha_selection, alpha_name){
    lapply(dataset_versions, function(train_set){
      net_x <- get(train_set) %>% 
        filter(feat_class%in%c("Good", "Bad")) %>%
        select(-feat_class, -feature) %>%
        data.matrix()
      net_y <- get(train_set) %>% 
        filter(feat_class%in%c("Good", "Bad")) %>%
        mutate(feat_class=feat_class=="Good") %>%
        pull(feat_class)
      net_model <- cv.glmnet(x=net_x, y=net_y, family="binomial", alpha=alpha_selection)

      lapply(dataset_versions, function(test_set){
        get(test_set) %>%
          select(-feat_class, -feature) %>%
          mutate(pred_prob=predict(object=net_model, newx = data.matrix(.), 
                                   type = "response")[,"lambda.1se"]) %>%
          mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
          mutate(feat_class=get(test_set)$feat_class) %>%
          select(feat_class, pred_class) %>%
          mutate(train_test=paste(train_set, test_set, sep = "-")) %>%
          mutate(train_test=str_remove_all(train_test, "_features")) %>%
          mutate(cross_type=ifelse(train_set==test_set, "Same", "Different"))
      }) %>% bind_rows()
    }) %>% bind_rows() %>% mutate(which_alpha=alpha_name)
  }) %>% bind_rows()
library(randomForest)
rf_cross_train_df <- lapply(dataset_versions, function(train_set){
  rf_x <- get(train_set) %>% 
    select(-medtop3_cor, -max_cor, -log_med_cor, -max_SNR) %>%
    select(med_SNR, med_cor, feat_class, feature) %>%
    filter(feat_class%in%c("Good", "Bad")) %>%
    select(-feat_class, -feature)
  rf_y <- get(train_set) %>% 
    filter(feat_class%in%c("Good", "Bad")) %>%
    pull(feat_class) %>%
    factor()
  rfmodel <- randomForest(rf_x, rf_y)

  lapply(dataset_versions, function(test_set){
    get(test_set) %>%
      select(-feat_class, -feature) %>%
      mutate(pred_class=predict(object=rfmodel, newdata = ., type = "response")) %>%
      mutate(feat_class=get(test_set)$feat_class) %>%
      select(feat_class, pred_class) %>%
      mutate(train_test=paste(train_set, test_set, sep = "-")) %>%
      mutate(train_test=str_remove_all(train_test, "_features")) %>%
      mutate(cross_type=ifelse(train_set==test_set, "Same", "Different"))
      }) %>% bind_rows()
    }) %>% bind_rows() %>%
  mutate(which_alpha="rf")

fdr_gpf_cross_train_df <- net_cross_train_df %>%
  rbind(rf_cross_train_df) %>%
    mutate(error_type=case_when(
    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
    feat_class=="Bad" & pred_class=="Good" ~ "FP",
    feat_class=="Good" & pred_class=="Bad" ~ "FN",
    feat_class=="Good" & pred_class=="Good" ~ "TP"
  )) %>%
  drop_na() %>%
  group_by(train_test, which_alpha, cross_type) %>%
  count(error_type) %>%
  ungroup() %>%
  mutate(cross_type=factor(cross_type, levels=c("Same", "Different"))) %>%
  pivot_wider(names_from=error_type, values_from = n, values_fill = 0) %>%
  mutate(FDR=FP/(FP+TP)) %>%
  mutate(GPF=TP/(FN+TP)) %>%
  mutate(cross_type_num=as.numeric(cross_type))
train_test_levels <- c("MS3000-MS3000", "FT2040-FT2040", "MS3000-FT2040", "FT2040-MS3000")
train_test_labels <- c("MESO-MESO", "Falkor-Falkor", "MESO-Falkor", "Falkor-MESO")
fdr_gpf_cross_train_gp <- fdr_gpf_cross_train_df %>%
  mutate(train_test=factor(train_test, levels=train_test_levels, labels=train_test_labels)) %>%
  mutate(which_alpha=factor(which_alpha, levels=c("ridge", "elastic", "lasso", "rf"),
                            labels=c("Ridge: α=0", "Elastic: α=0.5", "Lasso: α=1", "Random forest"))) %>%
  pivot_longer(c(FDR, GPF)) %>%
  ggplot() +
  geom_smooth(aes(x=cross_type_num, y=value, group=name), 
              method = "lm", formula="y~x", color="black", se=FALSE) +
  geom_point(aes(x=cross_type, y=value, color=train_test, 
                 fill=train_test, shape=name), size=5, alpha=0.8) +
  facet_grid(~which_alpha) +
  scale_x_discrete(labels = c("Same\ndataset\n", "Different\ndataset\n"),
                   breaks = c("Same", "Different")) +
  scale_color_manual(values = c("#0b505c", "#184d1f", "#0d8299", "#028e34"), 
                     breaks = train_test_labels,
                     aesthetics = c("color", "fill")) +
  scale_shape_manual(breaks = c("GPF", "FDR"), values = c("\u25B2", "\u25BC")) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_bw() +
  theme() +
  labs(x=NULL, y=NULL, color="Train-test", fill="Train-test", shape="Error type") +
  guides(shape=guide_legend(order=1))

ggsave(plot = fdr_gpf_cross_train_gp, 
       filename = "regregrf_fdrgpf_cross_train.png", 
       device = "png", path = "manuscript/supplement",
       width = 6.5, height = 3, units = "in", dpi = 300)
```

### Implications for biological conclusions

[Figure 7] shows the distribution of all mass features identified by XCMS and their categorization according to both the biochemical response to depth as well as the classifications assigned by the MS expert and the two-parameter model at a 50% likelihood threshold.

The abundance of noise peaks located at the center of the plot and therefore lacking much biological trend with depth validates the conventional heuristic of relying on downstream statistical tests to separate signal from noise (with the assumption that noise will be fixed across all samples and unaffected by environmental structuring of samples). However, it is also clear that there are both some good peaks that do not have trends with depth (and are therefore interesting for bucking the biomass signal) as well as a good number of bad peaks that do change across sample depths. These bad peaks that have biological signals appear to be frequently due to the tail of a peak being integrated as a separate molecular feature, which then retains the biological signal of the actual feature.

Most metabolites track with biomass and are consistently abundant through the surface and DCM but drop off at depth, as prior measurements of particulate carbon and beam attenuation have shown [@benedetto]. However, several groups of compounds diverge intriguingly from this bulk biomass trend, with some molecules found almost exclusively at the surface and others found to increase with depth rather than decrease. Several mass features that were found most exclusively at the surface were putatively identified as mycosporine-like amino acids (MAAs), which have been shown to have UV photoprotective properties.[@ref] Although we were unable to calculate their exact concentrations due to a lack of authentic standards, peak areas were approximately 100 to 1000 times higher at 15 meters than their values averaged between the DCM and 175 meter samples. The abundance of MAAs in this particular category indicates that there may be other MAA molecules with this depth pattern that have not yet been characterized also among these unknowns.

```{r clockplot BMIS, eval=!file.exists("manuscript/MS3000_norm_areas.csv")}
library(xcms)
file_data <- read_csv("made_data_MS3000/file_data.csv") %>%
  mutate(filename=basename(filename))
msnexp_filled <- readRDS("made_data_MS3000/msnexp_filled.rds")
peak_data_long <- msnexp_filled %>%
  chromPeaks() %>%
  as.data.frame() %>%
  rownames_to_column() %>%
  mutate(peakidx=row_number()) %>%
  select(sample, peakidx, into)
peak_data <- msnexp_filled %>%
  featureDefinitions() %>%
  as.data.frame() %>%
  filter(rtmed>30 & rtmed<1200) %>%
  select(peakidx) %>%
  rownames_to_column("feature") %>%
  unnest_longer(peakidx) %>%
  left_join(peak_data_long, by = join_by(peakidx)) %>%
  mutate(filename=basename(fileNames(msnexp_filled))[sample]) %>%
  complete(feature, filename, fill = list(into=0)) %>%
  left_join(file_data, by = join_by(filename)) %>%
  filter(str_detect(filename, "180821"))

source("manuscript/IS_integrations/is_combine.R")
IS_areas <- read_csv("manuscript/IS_integrations/all_IS.csv") %>%
  mutate(filename=paste0(filename, ".mzML")) %>%
  rename(area="IS_area") %>%
  filter(!str_detect(filename, "QC")) %>%
  pivot_wider(names_from=IS_name, values_from = "IS_area") %>%
  mutate("Inj_vol"=ifelse(str_detect(filename, "Poo.*Half"), 0.5, 1)) %>%
  pivot_longer(-filename, names_to = "IS_name", values_to = "IS_area")
pooled_IS_areas <- IS_areas %>%
  filter(str_detect(filename, "_Poo_"))

BMIS <- peak_data %>%
  filter(str_detect(filename, "_Poo_")) %>%
  full_join(pooled_IS_areas, multiple = "all", by = join_by(filename)) %>%
  group_by(feature, IS_name) %>%
  mutate(plot_area=into/IS_area*mean(IS_area)) %>%
  summarise(cv=sd(plot_area)/mean(plot_area)) %>%
  # left_join(init_cvs) %>%
  arrange(cv) %>%
  slice(1) %>%
  select(feature, IS_name)

BMISed_data <- peak_data %>%
  filter(samp_type=="Smp") %>%
  select(feature, filename, into) %>%
  left_join(BMIS, by = join_by(feature)) %>%
  left_join(IS_areas, by = join_by(filename, IS_name)) %>%
  group_by(feature) %>%
  mutate(norm_area=into/IS_area*mean(IS_area)) %>%
  select(feature, filename, norm_area)

norm_area_data <- BMISed_data %>%
  arrange(norm_area) %>%
  group_by(feature, filename) %>%
  slice(1) %>%
  ungroup()

if(any(is.na(norm_area_data$norm_area)))stop("Found NAs in norm_area!")
if(!all(table(norm_area_data$feature)==99))stop("Found not 99 samps in plotready!")
write_csv(norm_area_data, "manuscript/MS3000_norm_areas.csv")
```

```{r clockplot dataset munging, eval=!file.exists("manuscript/MS3000_clockplot_coords.csv")}
norm_area_data <- read_csv("manuscript/MS3000_norm_areas.csv")

both_min_model <- rbind(FT2040_features, MS3000_features) %>%
  select(feat_class, med_cor, med_SNR) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)
best_feats <- MS3000_features %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
  select(feature, feat_class, pred_class)
clockplot_data <- norm_area_data %>%
  right_join(best_feats, by = join_by(feature)) %>%
  left_join(file_data %>% select(filename, depth), by = join_by(filename))

kw_pvals_all <- clockplot_data %>%
  nest(data=-c(feature, pred_class)) %>%
  mutate(tidyaov=map(
    data, ~ broom::tidy(kruskal.test(.x$norm_area~.x$depth))), 
  ) %>%
  unnest(tidyaov) %>%
  select(-parameter, -method) %>%
  mutate(cor_pval_all=p.adjust(p.value, method="fdr")) %>%
  group_by(pred_class) %>%
  mutate(fdr_pval_byclass=p.adjust(p.value, method="fdr")) %>%
  ungroup()

resp_class_levels <- c("All equal", 
    "15m > DCM = 175m", "15m > DCM > 175m", "15m = DCM > 175m",
    "DCM > 15m > 175m", "DCM > 15m = 175m", "DCM > 175m > 15m", 
    "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m", 
    "Hard to tell")
kw_classes_all <- clockplot_data %>%
  group_by(feature) %>%
  mutate(norm_area=rank(norm_area)) %>%
  filter(var(norm_area)>0) %>%
  rstatix::dunn_test(norm_area~depth, p.adjust.method = "none") %>%
  mutate(contrast=paste(group2, group1, sep = "_")) %>%
  select(feature, contrast, est=statistic, pval=p.adj) %>%
  pivot_wider(names_from = contrast, values_from = c(pval, est)) %>%
  mutate(resp_class=case_when(
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m<0 & est_175m_DCM<0 ~ "15m > DCM > 175m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m<0 & est_175m_DCM<0 ~ "DCM > 15m > 175m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m>0 & est_175m_DCM>0 ~ "DCM < 15m < 175m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m>0 & est_175m_DCM>0 ~ "15m < DCM < 175m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_15m>0 & est_175m_DCM<0 ~ "DCM > 175m > 15m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_15m<0 & est_175m_DCM>0 ~ "DCM < 175m < 15m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM>0.05 & est_DCM_15m<0 & est_175m_15m<0 ~ "15m > DCM = 175m",
    pval_DCM_15m<0.05 & pval_175m_15m<0.05 & pval_175m_DCM>0.05 & est_DCM_15m>0 & est_175m_15m>0 ~ "15m < DCM = 175m",
    pval_DCM_15m>0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_175m_15m<0 & est_175m_DCM<0 ~ "15m = DCM > 175m",
    pval_DCM_15m>0.05 & pval_175m_15m<0.05 & pval_175m_DCM<0.05 & est_175m_15m>0 & est_175m_DCM>0 ~ "15m = DCM < 175m",
    pval_DCM_15m<0.05 & pval_175m_15m>0.05 & pval_175m_DCM<0.05 & est_DCM_15m>0 & est_175m_DCM<0 ~ "DCM > 15m = 175m",
    pval_DCM_15m<0.05 & pval_175m_15m>0.05 & pval_175m_DCM<0.05 & est_DCM_15m<0 & est_175m_DCM>0 ~ "DCM < 15m = 175m",
    pval_DCM_15m>0.05 & pval_175m_15m>0.05 & pval_175m_DCM>0.05 ~ "All equal",
    TRUE ~ "Hard to tell"
  )) %>%
  select(feature, resp_class) %>%
  mutate(resp_class=factor(resp_class, levels=resp_class_levels))

clockplot_xy_spots <- clockplot_data %>%
  group_by(feature) %>%
  mutate(norm_area=rank(norm_area)) %>%
  group_by(feature, depth) %>%
  summarise(mean_fold_change=mean(norm_area/50), .groups = "drop") %>% 
  # div by 50 bc 99 files and median 1:99 is 50
  pivot_wider(names_from=depth, values_from=mean_fold_change) %>%
  mutate(x=DCM-1) %>%           # cursed but works??
  mutate(y=`15m`+DCM/2-1.5) %>% # super fucking cursed
  select(feature, x, y)

clockplot_complete_df <- clockplot_data %>%
  distinct(feature, feat_class, pred_class) %>%
  left_join(clockplot_xy_spots, by = join_by(feature)) %>%
  left_join(kw_pvals_all %>% select(feature, fdr_pval_byclass), by = join_by(feature)) %>%
  left_join(kw_classes_all, by = join_by(feature)) %>%
  mutate(error_type=case_when(
    pred_class=="Good" & feat_class=="Good" ~ "TP",
    pred_class=="Good" & feat_class=="Bad" ~ "FP",
    pred_class=="Bad" & feat_class=="Good" ~ "FN",
    pred_class=="Bad" & feat_class=="Bad" ~ "TN"
  )) %>%
  mutate(error_type=factor(error_type, levels=c("TP", "FP", "FN", "TN"),
                           labels=c("Good peak found", "Bad peak included",
                                    "Good peak missed", "Bad peak avoided")))

write_csv(clockplot_complete_df, file = "manuscript/MS3000_clockplot_coords.csv")
```

```{r clockplot}
resp_class_levels <- c("All equal", 
    "15m > DCM = 175m", "15m > DCM > 175m", "15m = DCM > 175m",
    "DCM > 15m > 175m", "DCM > 15m = 175m", "DCM > 175m > 15m", 
    "15m < DCM = 175m", "15m < DCM < 175m", "15m = DCM < 175m", 
    "DCM < 15m < 175m", "DCM < 15m = 175m", "DCM < 175m < 15m", 
    "Hard to tell")
clockplot_complete_df <- read_csv(file = "manuscript/MS3000_clockplot_coords.csv") %>%
  mutate(resp_class=factor(resp_class, levels=resp_class_levels))

cmap_name <- "infinity"
cmap_url <- paste0("https://raw.githubusercontent.com/1313e/CMasher/master/",
                   "cmasher/colormaps/", cmap_name, "/", cmap_name, "_norm.txt")
cmtable <- read.table(cmap_url)
cmscale <- rgb(cmtable[,1], cmtable[,2], cmtable[,3])
color_scale <- c("grey50", head(cmscale[seq(1, length(cmscale), length.out=13)], 12), "grey80")
names(color_scale) <- resp_class_levels

clock_scale <- 0.66
axis_xends <- c(-1/2, 1.1, -1/2)*clock_scale
axis_yends <- c(sqrt(3)/2, 0, -sqrt(3)/2)*clock_scale
axis_y_dodge <- c(2, 2, -2)*clock_scale/25

clockplot_base_gp <- ggplot() +
  annotate("segment", x=c(0,0,0), y=c(0,0,0), 
           xend=axis_xends, yend=axis_yends,
           arrow=arrow(length=unit(0.02, "npc"), type = "closed")) +
  annotate("segment", x=c(0,0,0), y=c(0,0,0),
           xend=-axis_xends, yend=-axis_yends,
           lty = 2) +
  annotate("label", x=axis_xends, y=axis_yends+axis_y_dodge,
           label=c("15m axis", "DCM axis", "175m axis"), 
           fill="#FFFFFFAA", label.size=0) +
  coord_fixed(xlim = c(-0.65, 1.3)*clock_scale, 
              ylim = c(-1, 1)*clock_scale, 
              expand=TRUE) +
  scale_color_manual(values = color_scale, breaks = names(color_scale), drop=FALSE) +
  theme_void() +
  theme(legend.position = "right", 
        legend.title = element_blank(),
        plot.background = element_rect(fill="white", color="white")) +
  guides(colour = guide_legend(override.aes = list(size=4, alpha=1), drop=FALSE)) +
  aes(x=x, y=y, color=resp_class, label=feature)

clockplot_complete_gp <- clockplot_base_gp +
  geom_point(aes(shape=error_type), alpha=0.5, size=2,
             data = clockplot_complete_df) +
  scale_shape_manual(breaks = c("Good peak found", "Bad peak included",
                                "Good peak missed", "Bad peak avoided"), 
                     values = c(19, 3, 4, 1))

ggsave(plot = clockplot_complete_gp, 
       filename = "clockplot_complete.png", 
       device = "png", path = "manuscript/figures",
       width = 6.5, height = 5, units = "in", dpi = 300)
```

![Figure 7: Plot of metabolite response to depth shown across three axes using the rank-normalized median value at each depth as the coordinate for that axis. Each molecular feature corresponds to a point in the plot, and their position on the plot describes the shape of their depth profile. Compounds aligning with the 15m axis correspond to compounds with most of their abundance found in the surface ocean; points far to the right side correspond to compounds that are found only at the deep chlorophyll maximum; points found at the bottom of the plot are those compounds that increased more or less linearly with depth.](figures/clockplot_complete.png){width=90%}

### 2023-05-02 manuscript/outline.Rmd
Adding regularized regression discussion and clockplot code

  - Clockplot and univariate statistics
    - Report proportions of MFs in each category?
    - Many "bad" still have strong biology trend, not a sufficient measure on its own
    - Missing out on some good peaks (far left clockplot)
    - Do the FDR corrected p-value comparisons with more/less peaks by threshold

>>>>

  - Clockplot and univariate statistics
    - Report proportions of MFs in each category?
    - Big central thing where all the noise ended up
    - Many "bad" still have strong biology trend, not a sufficient measure on its own
    - Missing out on some good peaks (far left clockplot)
    - Do the FDR corrected p-value comparisons with more/less peaks by threshold
